A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-08-31 bioc-workflow.Rmd
Updated naming of msnsets, added pData

To use the `readMSnSet2` function, as a minimum one must specify the
file path to the data and which columns of the spreadsheet contain
quantitation data. The `getEcols` function exists to help users
identify which columns of the spreadsheet contain the
quantitation data.

We first locate the the spreadsheet of E14TG2a data distributed with
the `r Biocexptpkg("pRolocdata")` data using `system.file`, that
locates the `extdata` in the `r Biocexptpkg("pRolocdata")` package on
the hard drive and `dir`, that displays the full path to the file
matching the `hyperLOPIT-SIData-ms3-rep12-intersect.csv` pattern,
which corresponds the the file of interest. In the last line, we print
the file name (not the full path, which will vary from computer to
computer).

@@

As briefly mentioned above, the quantitation data is stored in the
`exprs` slot of the `MSnSet` and can be accessed by
`exprs(hl)`. Below, we examine the quantitative information for
first 5 proteins. It is also possible to access specific rows and
columns by naming the proteins and fractions of interest.

@@

```{r pdata}
pData(hl)$replicate <- rep(1:2, each = 10)
pData(hl)$tag <- sub("\\.1$", "", sub("^X", "", sampleNames(hl)))
pData(hl)
```

@@

There are two aspects related to data normalisation that are relevant
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between fractions without
affecting biological variability (i.e. the shape of the quantitative
profiles) too much, as generally focused on. This normalisation will
depend on the underlying quantitative technology and the experimental
design, and will not be addressed in this workflow. In addition, and
more specific to spatial proteomics data, it is fundamental to scale
all the organelle-specific profiles into a same intensity interval
(typically 0 and 1) by, for example, diving each intensity by the sum
of the intensities for that quantitative feature, as illustrated
below.

@@

Before combination, the two replicates in the `hl` data that we
read into R were separately normalised by sum across the 10 channels
for each replicate respectively. We can verify this by summing each
rows for each replicate:

```{r normcheck}
summary(rowSums(exprs(hl[, hl$replicate == 1])))
summary(rowSums(exprs(hl[, hl$replicate == 2])))
```
We see that some features do not add up exactly to 1 due to rounding
errors after exporting to intermediate files. These small deviations
do not bear any consequences here.

@@

The spreadsheet that was used to create the `hl` MSnSet
included two replicates. We also provide individual replicates in the
`r Biocexptpkg("pRolocdata")` package. Below, we show how to combine
`MSnSet` objects and, subsequently, how to filter and handle missing
values. We start by loading the experimental data package and the
equivalent replicates using the `load` function.

@@

```{r fvarnames2}
fvarLabels(hyperLOPIT2015ms3r1)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r1)[3:5], 1)
fvarLabels(hyperLOPIT2015ms3r2)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r2)[3:5], 2)
fData(hyperLOPIT2015ms3r2) <- fData(hyperLOPIT2015ms3r2)[3:5]
```

@@

```{r combine}
hyperLOPIT <- combine(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)
hyperLOPIT
```

@@

```{r namap}
image2(is.na(hyperLOPIT), col = c("black", "white"),
       main = "Missing values (white cells) after combining replicates")
```

@@

```{r filterNA}
hyperLOPIT <- filterNA(hyperLOPIT)
hyperLOPIT
```

@@

```{r qcplot, message=FALSE, warning=FALSE}
library("pRoloc")
plot2D(hyperLOPIT, fcol = NULL)
```

@@

**TODO** There are other dimensionality reduction methods available in
the `plot2D` function, which can be parametrised with the `method`
argument.  Advantages, disadvantages? See
[this](https://gist.github.com/mikelove/74bbf5c41010ae1dc94281cface90d32)
very interesting gist with intriguing tSNE examples. This might be
better to come after markers.

@@

```{r qcbx, fig.width = 12}
par(mfrow = c(1, 2))
o <- order(hyperLOPIT$Iodixonal.Density)
plotDist(hyperLOPIT[, o], pcol = "#00000010")
lines(colMeans(exprs(hyperLOPIT[, o])), col = "red", type = "b")
boxplot(exprs(hyperLOPIT[, o]), las = 2)
```

@@

```{r plotmarkers, fig.width = 12}
par(mfrow = c(1, 2))
plot2D(hl, fcol = "markers", main = "pRolocmarkers for mouse")
addLegend(hl, fcol = "markers", where = "bottomleft", cex = .7)
plot2D(hl, dims = c(1, 7), main = "Marker resolution along PC 1 and 7")
```

@@

```{r plotDist}
hlo <- hyperLOPIT[, order(hyperLOPIT$Iodixonal.Density)]
plotDist(hlo[fData(hlo)$markers == "Mitochondrion", ],
         pcol = "purple", fractions = "Fraction.No")
title(main = "Marker occupancy profiles along the gradient")
matlines(t(exprs(hlo[fData(hlo)$markers == "Peroxisome", ])),
         lty = 1, col = "darkblue", type = "l")
legend("topleft", c("Mitochondrion", "Peroxisome"),
       lty = 1, col = c("purple", "blue"), bty = "n")
```

@@

```{r pdtab, echo=FALSE, eval = FALSE}
i <- hyperLOPIT$Fraction.No %in% 7:10
pData(hyperLOPIT)[i, -3]
```

@@

```{r plot2Drep, fig.width = 12}
par(mfrow = c(1, 2))
plot2D(hl[, hl$replicate == 1], main = "Replicate 1")
plot2D(hl[, hl$replicate == 2], main = "Replicate 2",
       mirrorX = TRUE)
```

@@

```{r}
i1 <- exprs(hl[, hl$replicate == 1])
i2 <- exprs(hl[, hl$replicate == 2])
pearcor <- cor(t(i1), t(i2))
pearcor0 <- cor(t(i1[, seq(1, 10, 2)]), t(i2[, seq(1, 10, 2)]))
mean(diag(pearcor))
mean(diag(pearcor0))
```

>>>>

To use the `readMSnSet2` function, as a minimum one must specify the
file path to the data and which columns of the spreadsheet contain
quantitation data. The `getEcols` function exists to help users
identify which columns of the spreadsheet contain the
quantitation data. In the last line of the code chunk below, we print
the file name (not the full path, which will vary from computer to
computer).

@@

Below, we examine the quantitative information for
first 5 proteins. It is also possible to access specific rows and
columns by naming the proteins and fractions of interest.

@@

```{r pdata}
pData(hl)$Replicate <- rep(1:2, each = 10)
pData(hl)$Tag <- sub("\\.1$", "", sub("^X", "", sampleNames(hl)))
pData(hl)
```

Sample meta-data distributed with
the `r Biocexptpkg("pRolocdata")` and as above with the quantitation
data file is located in the `extdata` in the `r Biocexptpkg("pRolocdata")` 
package on the hard drive. In the code chunk below we again use the `dir` 
function to locate the filepath to the meta-data `csv` file and then read 
it into R using `read.csv` and append the meta-data to the `pData` slot. 
We add information on the gradient fractions used and the associated 
subcellular fraction densities in % w/v Iodixanol for each replicate.

```{r morepdata}
expinfo <- dir(extdatadir, full.names = TRUE,
               pattern = "hyperLOPIT-SIData-fraction-info.csv")

fracinfo <- read.csv(expinfo, row.names=1, skip = 2, 
                     header = FALSE, stringsAsFactors = FALSE)

pData(hl)$Gradient.Fraction <- c(fracinfo[, 1], fracinfo[, 2])
pData(hl)$Iodixonal.Density <- c(fracinfo[, 4], fracinfo[, 5])
pData(hl)
```

@@

There are two aspects related to data normalisation that are relevant
to spatial proteomics data processing. The first one focuses on
reducing purely technical variation between fractions without
affecting biological variability (i.e. the shape of the quantitative
profiles) too much, as generally focused on. This normalisation will
depend on the underlying quantitative technology and the experimental
design, and will not be addressed in this workflow. In addition, and
more specific to spatial proteomics data, it is fundamental to scale
all the organelle-specific profiles into a same intensity interval
(typically 0 and 1) by, for example, diving each intensity by the sum
of the intensities for that quantitative feature. This is not necessary 
in this example as the intensities for each replicate have already 
been re-scaled to 1 in Proteome Discoverer. However, if one wanted
to do this they would execute the `normalise` function as demonstrated 
in the below code chunk.

@@

As previously mentioned, before combination, the two replicates in 
the `hl` data that we read into R were separately normalised 
by sum (i.e. to 1) across the 10 channels for each replicate 
respectively. We can verify this by summing each rows for each replicate:

```{r normcheck}
summary(rowSums(exprs(hl[, hl$replicate == 1])))
summary(rowSums(exprs(hl[, hl$replicate == 2])))
```

We see that some features do not add up exactly to 1 due to rounding
errors after exporting to intermediate files. These small deviations
do not bear any consequences here.

@@

The spreadsheet that was used to create the `hl` MSnSet
included two replicates. We also provide individual replicates in the
`r Biocexptpkg("pRolocdata")` package. Below, we show how to combine
`MSnSet` objects and, subsequently, how to filter and handle missing
values. We start by loading the experimental data package and the
equivalent replicates using the `data` function.

@@

```{r fvarnames2}
fvarLabels(hyperLOPIT2015ms3r1)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r1)[3:5], 1)
fvarLabels(hyperLOPIT2015ms3r2)[3:5] <- paste0(fvarLabels(hyperLOPIT2015ms3r2)[3:5], 2)
fData(hyperLOPIT2015ms3r1) <- fData(hyperLOPIT2015ms3r1)[1:5] 
fData(hyperLOPIT2015ms3r2) <- fData(hyperLOPIT2015ms3r2)[3:5] 
```

@@

```{r combine}
combined <- combine(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)
combined
```

@@

```{r namap}
image2(is.na(combined), col = c("black", "white"),
       main = "Missing values (white cells) after combining replicates")
```

@@

```{r filterNA}
combined <- filterNA(combined)
combined
```

@@

```{r qcplot, message=FALSE, warning=FALSE}
library("pRoloc")
plot2D(hl, fcol = NULL, col = "black")
```

@@

**TODO** There are other dimensionality reduction methods available in
the `plot2D` function, which can be parametrised with the `method`
argument.  Advantages, disadvantages? See
[this](https://gist.github.com/mikelove/74bbf5c41010ae1dc94281cface90d32)
very interesting gist with intriguing tSNE examples. <!--This might be
better to come after markers.-->

@@

```{r qcbx, fig.width = 12}
par(mfrow = c(1, 2))
o <- order(hl$Iodixonal.Density)
plotDist(hl[, o], pcol = "#00000010")
lines(colMeans(exprs(hl[, o])), col = "red", type = "b")
boxplot(exprs(hl[, o]), las = 2)
```

@@

```{r plotmarkers, fig.width = 12}
par(mfrow = c(1, 2))
plot2D(hl, main = "pRolocmarkers for mouse")
addLegend(hl, cex = .7)
plot2D(hl, dims = c(1, 7), main = "Marker resolution along PC 1 and 7")
```

@@

```{r plotDist}
hlo <- hl[, order(hl$Iodixonal.Density)]
plotDist(hlo[fData(hlo)$markers == "Mitochondrion", ],
         pcol = "purple", fractions = "Fraction.No")
title(main = "Marker occupancy profiles along the gradient")
matlines(t(exprs(hlo[fData(hlo)$markers == "Peroxisome", ])),
         lty = 1, col = "darkblue", type = "l")
legend("topleft", c("Mitochondrion", "Peroxisome"),
       lty = 1, col = c("purple", "blue"), bty = "n")
```

@@

```{r pdtab, echo=FALSE, eval = FALSE}
i <- c(2, 3, 4, 13, 14)
pData(hl)[i, -3]
```

@@

```{r plot2Drep, fig.width = 12}
par(mfrow = c(1, 2))
plot2D(hl[, hl$Replicate == 1], main = "Replicate 1")
plot2D(hl[, hl$Replicate == 2], main = "Replicate 2",
       mirrorX = TRUE)
```

@@

```{r pearson}
i1 <- exprs(hl[, hl$Replicate == 1])
i2 <- exprs(hl[, hl$Replicate == 2])
pearcor <- cor(t(i1), t(i2))
pearcor0 <- cor(t(i1[, seq(1, 10, 2)]), t(i2[, seq(1, 10, 2)]))
mean(diag(pearcor))
mean(diag(pearcor0))
```
